% =============================================================================
% DRAFT section for main.tex -- "Source-free protocol: fair by construction".
% Numbers are macros; wire them through scripts/make_arc3_assets.py (do NOT hand-edit numbers.tex).
% Macro names are letters-only (no digits) per the house LaTeX rule.
% Suggested placement: immediately AFTER \section{The ARC-AGI-3 setting} and BEFORE the recipe,
% so the protocol is fixed before any result is reported. Cross-ref \label{sec:sourcefree}.
% =============================================================================
\section{Source-free protocol: fair by construction}
\label{sec:sourcefree}

\textbf{The concern.} The downloadable ARC-AGI-3 environment ships each game's \emph{Python source}.
An agent running in the same process and working directory can therefore read it---either by
\texttt{importlib}-loading \texttt{<game>.py} or by \texttt{inspect.getsource(env.\_game)}---and recover
the win condition directly. That trivialises the two capabilities the benchmark is meant to test:
``discover the dynamics \emph{by acting}'' and ``\emph{reason} the win condition.'' Reading the source
is reading the answer key, and it breaks comparability to the live leaderboard agents, which cannot.

\textbf{We audited our own initial runs and found the leak.} A file-access audit of the first
full-game push shows \ArcSourceUsedFull{} of the \ArcSourceAssistedFull{} full-game solves consulted
the provided source (an \texttt{importlib} load of \texttt{<game>.py}, or hand reverse-engineering with
line citations). We report that number plainly rather than bury it: the \emph{source-assisted}
full-game total is \ArcSourceAssistedFull/\ArcUnionGames{}, but it is not a fair-agent result.

\textbf{The fix is a sandbox that makes source access impossible, not merely discouraged.} Prompting
the agent ``do not read the source'' is unverifiable and, empirically, was not obeyed. Instead we
re-run every game under a protocol that is \emph{source-free by construction} (three guarantees,
each independently checkable):
\begin{enumerate}\setlength{\itemsep}{1pt}
\item \textbf{Process isolation.} The real \texttt{arc\_agi} game runs in a separate \emph{worker}
process whose working directory holds the downloaded source. The agent imports only
\texttt{SandboxGame}, a thin pipe client exposing \emph{only} \{\texttt{frame}, \texttt{levels},
\texttt{win}, \texttt{avail}, \texttt{done}\}. No game object exists in the agent's process, so
\texttt{inspect.getsource} has no target.
\item \textbf{Clean filesystem.} The agent's working directory contains only the client harness---%
\emph{no} \texttt{environment\_files}---and the agent interpreter cannot \texttt{import arc\_agi}. There
is no \texttt{<game>.py} to load.
\item \textbf{Audit-gated banking.} Every solve's working directory is audited
(\texttt{scripts/audit\_sandbox.py}: no \texttt{environment\_files}, and no script referencing
\texttt{getsource} / \texttt{spec\_from\_file\_location} / \texttt{arc\_agi}); only audit-clean,
replay-verified solves are banked into \texttt{arc3\_fullgame\_sourcefree.json}.
\end{enumerate}
The agent thus \emph{must} gather \((s,a,s')\) transitions through the pipe, synthesise
\texttt{predict(frame,action)} from observation, and reason the win from the frames it sees---the
intended task. A smoke test confirms the guarantees end-to-end: a \texttt{numpy}-only agent process
(which cannot import \texttt{arc\_agi}) drives the game, its working tree contains no game source, and
no game object is reachable from the client (the worker is a distinct OS process).

\textbf{Result (source-free).} Under this protocol the executable-world-model recipe completes
\textbf{\ArcSourceFreeFull/\ArcUnionGames{} games in full} (\ArcSourceFreeLevels/\ArcTotalLevels{}
levels), every solve audit-clean and replay-verified. This is the number we compare to the prior state
of the art (\ArcBaselineFull/\ArcUnionGames~\cite{baseline1_ewm}): \ArcSourceFreeVsSota. We report the
source-assisted \ArcSourceAssistedFull/\ArcUnionGames{} separately and clearly labelled, never as the
headline.

\textbf{The boundary is the diagnosis.} The games that resist \emph{source-free} solving are the deep
\emph{procedural} ones---precisely where the win must be \emph{reasoned} from observation rather than
\emph{read}. The source-free frontier is therefore not an artifact of the sandbox but a direct
measurement of the goal-as-procedure wall (\S\ref{sec:solvewall}): removing the answer key leaves
exactly the levels whose win is an ordered protocol no observation-only score expresses.
